Creating figures for the paper

### 2024-08-07 infercnv/reports/paper_figures.Rmd
trying different lolliplot methods

@@

```{r}
lh22_dir <- "/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/"

@@

# load the muts and convert to list
load(file.path(lh22_dir, "Figures/Figure_2/LCM_NanoSeq_Overlap/GOI",
               "pileup_muts.rda"))
muts <- list(exome = exome_muts,
             LCM = LCM_muts,
             targeted = nano_muts,
             pileup = pileup_muts)

# lollipopPlot

# genes of interest
goi <- "NF1"

# get annotations from dndscv output
mut_annots <-
  muts$exome %>%
  tibble::as_tibble() %>%
  dplyr::select(chr, pos, gene, ref, mut, type, tri, impact, aachange)

muts$targeted %>%
  tibble::as_tibble() %>%
  dplyr::select()
```

>>>>

# Datasets

## LCM NanoSeq

@@

```{r}
# add library path
.libPaths(c("/lustre/scratch125/casm/team268im/at31/.R/lib/4.2.2/", .libPaths()))

# libraries
library(magrittr)
library(GenomicRanges)
library(trackViewer)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)  # package for gene model
library(org.Hs.eg.db)                       # package for gene name
library(biomaRt)
library(httr)                               # package for REST API

# dirs
lh22_dir <- "/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/"

@@

# load the muts (targeted mutations in the nano_muts object)
load(file.path(lh22_dir, "Figures/Figure_2/LCM_NanoSeq_Overlap/GOI",
               "pileup_muts.rda"))
nano_muts <-
  nano_muts %>%
  tibble::as_tibble()

# get mutations, convert to granges
gr_muts_ls <-
  nano_muts %>%
  tibble::as_tibble() %>%
  dplyr::add_count(mut_site) %>%
  {
    GenomicRanges::GRanges(
      seqnames = paste0("chr", .$chr),
      ranges = IRanges::IRanges(start = .$pos,
                                width = 1,
                                names = .$mut_site),
      strand = ifelse(.$strand == 1, "+", "-"),
      gene = .$gene, score = .$n)
  } %>%
  {split(., .$gene)}

# get the refcds
data("refcds_hg19", package = "dndscv")
names(RefCDS) <- RefCDS %>% purrr::map(~ .$gene_name)

# write urls for proteinpaint

# write custom input file for proteinpaint
nano_muts %>%
  dplyr::filter(gene == "NF1") %>%
  dplyr::transmute(
    aachange, 
    aapos = as.numeric(stringr::str_sub(aachange, 2, -2)),
    impact = dplyr::case_when(impact == "Missense" ~ "M",
                              impact == "Nonsense" ~ "N",
                              impact == "Synonymous" ~ "S",
                              impact == "no-SNV" & type == "del" ~ "D",
                              impact == "no-SNV" & type == "ins" ~ "I",
                              TRUE ~ NA),
    donor) %>%
  dplyr::filter(!is.na(aapos)) %>%
  readr::write_csv("protein_paint.csv")

# write custom input file for mutationmapper
nano_muts %>%
  dplyr::filter(gene == "NF1") %>%
  dplyr::transmute(
    Chromosome = chr, Start_Position = pos - 1, End_Position = pos,
    Reference_Allele = ref, Variant_Allele = mut, Hugo_Symbol = gene
  ) %>%
  readr::write_tsv("mutation_mapper.tsv")

# get the gene
g <- "NF1"

# retrieve refcds element for gene
x <- RefCDS[[g]]

# get the position of each base along the exons
# RefCDS is 1-based - must convert to 0-based for calculations, then
# convert back to 1-based for plotting
exons <-
  tibble::tibble(start = x$intervals_cds[, 1] - 1,
                 end = x$intervals_cds[, 2]) %>%
  dplyr::mutate(dplyr::across(everything(), 
                              ~ .x - (x$intervals_cds[1, 1] - 1),
                              .names = "rel_{.col}"),
                exon = dplyr::row_number(),
                exon_length = end - start,
                cds_end = cumsum(exon_length),
                cds_start = dplyr::lag(cds_end, default = 0),
                # convert starts to 1-based for plotting
                dplyr::across(dplyr::contains("start"),
                              ~ .x + 1))

# expand to get every position
exons_exp <-
  exons %>%
  dplyr::group_by(rel_start, rel_end, exon, exon_length) %>%
  dplyr::reframe(pos = start:end, cds_pos = cds_start:cds_end)

# load interpro entries
interpro2go <-
  readLines("data/interpro/interpro2go.txt") %>%
  tibble::as_tibble() %>%
  dplyr::filter(!grepl("^!", value)) %>%
  tidyr::separate_wider_delim("value", delim = " ",
                              names = c("interpro_accession", "rest"),
                              too_many = "merge") %>%
  tidyr::separate(rest, sep = " > GO:| ; GO:",
                  into = c("domain_name", "GO_name", "GO_id")) %>%
  dplyr::mutate(interpro_accession = gsub("InterPro:", "", interpro_accession))
interpro <-
  readr::read_tsv("data/interpro/entry.list") %>%
  dplyr::rename(interpro_accession = ENTRY_AC)

# retrieve protein domain annotations
library(ensembldb)
library(EnsDb.Hsapiens.v75)
edb <- EnsDb.Hsapiens.v75
domains <-
  transcripts(edb, filter = GeneNameFilter(g),
              columns = c("protein_id", "tx_biotype", "uniprot_id",
                          "protein_domain_id", "protein_domain_source",
                          "prot_dom_start", "prot_dom_end",
                          "interpro_accession"))
domains_exp <-
  txs %>%
  dplyr::as_tibble() %>%
  dplyr::left_join(interpro2go) %>%
  dplyr::filter(tx_biotype == "protein_coding",
                !is.na(interpro_accession), !is.na(domain_name)) %>%
  dplyr::mutate(dom_start = start + prot_dom_start,
                dom_end = start + prot_dom_end) %>%
  dplyr::group_by(tx_id, protein_id, protein_domain_id, protein_domain_source,
                  prot_dom_start, prot_dom_end, interpro_accession,
                  domain_name, GO_name) %>%
  dplyr::reframe(pos = dom_start:dom_end)

# combine all exon and domain annotations
annots <-
  exons_exp %>%
  dplyr::full_join(domains_exp)

# add exon annots to muts
impact_palette <-
  c("Synonymous" = "#b3b3b3",
    "Missense" = "#5f9ea0",
    "Nonsense" = "#68218a")
p_dat <-
  nano_muts %>%
  dplyr::group_by(chr, pos, impact) %>%
  dplyr::count() %>%
  dplyr::right_join(annots) %>%
  dplyr::mutate(alternate = ifelse(exon %% 2 == 0, "e", "o"),
                impact = factor(impact, levels = names(impact_palette)))

# plot
pdf("test.pdf", width = 30)
# mutations
p_dat %>%
  ggplot(aes(x = cds_pos, y = n, fill = impact, colour = impact)) +
  geom_col(width = 1) +
  geom_point() +
  theme_classic() +
  scale_fill_manual(values = impact_palette,
                    na.value = "white", na.translate = FALSE) +
  scale_colour_manual(values = impact_palette,
                      na.value = "white", na.translate = FALSE)
# exon bounds
p_dat %>%
  dplyr::group_by(exon, alternate) %>%
  dplyr::summarise(xmin = min(cds_pos), xmax = max(cds_pos)) %>%
  ggplot(aes(xmin = xmin, xmax = xmax, ymin = -1, ymax = 1,
             fill = alternate)) +
  geom_hline(yintercept = 0, colour = "black") +
  geom_rect(colour = "black") +
  theme_void() +
  theme(axis.text.x = element_text()) +
  scale_fill_manual(values = c("e" = "#e3b3b2",
                               "o" = "#cd6f6e"))
# protein domains
p_dat %>%
  dplyr::filter(!is.na(protein_domain_id)) %>%
  dplyr::group_by(tx_id, domain_name, GO_name, protein_domain_id) %>%
  dplyr::summarise(xmin = min(cds_pos), xmax = max(cds_pos)) %>%
  ggplot(aes(xmin = xmin, xmax = xmax, ymin = -1, ymax = 1,
             fill = GO_name)) +
  geom_hline(yintercept = 0, colour = "black") +
  geom_rect(colour = "black") +
  theme_void() +
  scale_fill_discrete(na.value = "white", na.translate = FALSE) +
  facet_grid(tx_id ~ .)

dev.off()

# get gene coords (GRCh37)
ensembl <- useEnsembl(biomart = "ensembl", dataset = "hsapiens_gene_ensembl",
                      GRCh = 37)
bm <-
  getBM(attributes = c("chromosome_name", "start_position",
                       "end_position", "strand", "hgnc_symbol"),
        filters = c("hgnc_symbol"),
        values = list(names(gr_muts_ls)),
        mart = ensembl)
gr_genes <-
  GenomicRanges::GRanges(
    seqnames = paste0("chr", bm$chromosome_name),
    ranges = IRanges::IRanges(start = bm$start_position,
                              end = bm$end_position,
                              names = bm$hgnc_symbol),
    strand = ifelse(bm$strand == 1, "+", "-"),
    gene = bm$hgnc_symbol)
gr_genes_ls <- split(gr_genes, gr_genes$gene)

# get transcripts per gene
nf1_trs <-
  geneModelFromTxdb(TxDb.Hsapiens.UCSC.hg19.knownGene,
                    org.Hs.eg.db, gr = gr_genes_ls$NF1) %>%
  purrr::map(~ .$dat) %>%
  {Reduce(c, .)} %>%
  subset(symbol == "NF1" & feature == "CDS")
nf1_trs_fill <-
  tibble::tibble(transcript = unique(nf1_trs$transcript),
                 fill = rainbow(dplyr::n_distinct(nf1_trs$transcript))) %>%
  tibble::column_to_rownames("transcript")
nf1_trs$fill <-  nf1_trs_fill[nf1_trs$transcript, ]
nf1_trs$featureLayerID <- nf1_trs$transcript
names(nf1_trs) <- nf1_trs$transcript

# get protein domains
APIurl <- "https://www.ebi.ac.uk/proteins/api/" # base URL of the API
taxid <- "9606" # human tax ID
gene <- "NF1" # target gene
orgDB <- "org.Hs.eg.db" # org database to get the uniprot accession id
eid <- mget(gene, get(sub(".db", "SYMBOL2EG", orgDB)))[[1]]
chr <- mget(eid, get(sub(".db", "CHR", orgDB)))[[1]]
accession <- unlist(lapply(eid, function(.ele){
  mget(.ele, get(sub(".db", "UNIPROT", orgDB)))
  }))
featureURL <- paste0(APIurl,
                      "features?offset=0&size=-1&reviewed=true",
                      "&types=DNA_BIND%2CMOTIF%2CDOMAIN",
                      "&taxid=", taxid,
                      "&accession=", paste(accession, collapse = "%2C"))
response <- GET(featureURL)
content <- httr::content(response)[[1]]
acc <- content$accession
sequence <- content$sequence
gr <- GRanges(chr, IRanges(1, nchar(sequence)))
domains <- do.call(rbind, content$features)
domains <- GRanges(chr, IRanges(as.numeric(domains[, "begin"]),
                                as.numeric(domains[, "end"]),
                                names = domains[, "description"]))
names(domains)[1] <- "DNA_BIND" ## this is hard coding.
domains$fill <- 1 + seq_along(domains)
domains$height <- 0.04
## GET variations. This part can be replaced by user-defined data.
variationURL <- paste0(APIurl,
                        "variation?offset=0&size=-1",
                        "&sourcetype=uniprot&dbtype=dbSNP",
                        "&taxid=", taxid,
                        "&accession=", acc)
response <- GET(variationURL)
content <- httr::content(response)[[1]]
keep <- sapply(content$features, function(.ele) length(.ele$evidences)>2 && # filter the data by at least 2 evidences
                  !grepl("Unclassified", .ele$clinicalSignificances)) # filter the data by classified clinical significances.
nkeep <- c("wildType", "alternativeSequence", "begin", "end",
            "somaticStatus", "consequenceType", "score")
content$features <- lapply(content$features[keep], function(.ele){
  .ele$score <- length(.ele$evidences)
  unlist(.ele[nkeep]) 
})
variation <- do.call(rbind, content$features)
variation <-
  GRanges(chr,
          IRanges(as.numeric(variation[, "begin"]),
                  width = 1,
                  names = paste0(variation[, "wildType"],
                                  variation[, "begin"],
                                  variation[, "alternativeSequence"])),
          score = as.numeric(variation[, "score"]),
          color = as.numeric(factor(variation[, "consequenceType"]))+1)
variation$label.parameter.gp <- gpar(cex=.5)
seqlevels(variation) <- paste0("chr", seqlevels(variation))
seqlevels(domains) <- paste0("chr", seqlevels(domains))

domains %>%
  tibble::as_tibble()

pdf("test.pdf")
lolliplot(gr_muts_ls$NF1, domains, nf1_trs)
dev.off()

# plot transcripts per gene
pdf("test.pdf", width = 20)
lolliplot(gr_muts_ls$NF1, nf1_trs)
dev.off()
```

```{r}
load(file="/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/Figures/Figure_1/GOI.rda")
goi <- list(exome_genes = exome_genes, goi_exome = goi_exome, 
            goi_targeted = goi_targeted, targeted_genes = targeted_genes)

nanoseqVarsPath <- "nano_muts.tsv"
```
